Cybersecurity Impacts of the Covid-19 Pandemic in Italy

### 2021-03-08 article.tex
Added email
Updated pictures
Added numbers of responses

% Institutes for affiliations are also joined by \and,
\institute{
  Associazione Italiana Professionisti della Sicurezza Informatica (AIPSI), Milano, Italy
\and
  Italian Information System Security Association International (ISSA)
\and
  Digital Attacks Observatory (OAD) Team, Italy
\and
   University of Verona, Verona, Italy\\
   \email{\{luca.olivieri,fausto.spoto\}@univr.it}\\
 }

@@

The \textit{Digital Attacks Observatory} (OAD)~\cite{oadweb} is the only independent online survey in Italy about intentional digital attacks on IT systems of companies and public
bodies. It is implemented in collaboration with the Italian postal and telecommunications police.
The last OAD 2020 was the twelfth year of consecutive surveys on cyberattacks in Italy.
The OAD survey is not directed at a predefined set of respondents, but allows potential interested
parties to have full and free access to an online questionnaire, in a totally anonymous way. 
Thanks to the number of responses collected and their balanced distribution between companies and public bodies of different size and belonging to different economic
sectors, we can provide a specific picture of the cyberattacks in Italy.

@@

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{pictures/fig3.png}
  \caption{Comparison between cyberattacks occurred in 2019 and in the first quarter of 2020,
    wrt.\ to the organization dimension~\cite{oad20}. The dotted line links
    the unreported attacks and shows that they decrease while moving from tiny to big organizations,
    in terms of employees.}
  \label{fig:3}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{pictures/fig2.png}
  \caption{Trend of the reported attacks from the beginning of the OAD survey
    in 2008~\cite{oad20}. This comparison is only valid as a trend and not as a statistical
    measure, since the respondents are different from year to year. The dotted red line shows
    successive waves: after a relative pick of attacks, companies react and strengthen
    their cybersecurity measures, so that, the following year, the trend dicreases.}
  \label{fig:2}
\end{figure}

The chart in Fig.~\ref{fig:2} analyzes the trend of cyberattaks.
The figure shows that, from 2007 to 2015, in average, about 40\% of the respondents
reported cyberattacks (leftmost full red line). Since 2015, the trend increased to 45\%
of the respondents (rightmost full red line). 
In 2018, OAD reached the pick of reported attacks and, for the first time, the percent of
reported attacks became larger than that percent of companies that did not report any attack.
After the pick 
in 2018, year 2019 has a lower figure and the same occurs in the first quarter 2020.
The trend for full 2020 will be discovered with the currently ongoing OAD survey.
This view, over several years, does not show significant changes, not even inside
the pandemic period, but shows only a slight increase in cyberattacks 
in the recent years, not necessarily attributable to the pandemic crisis.

@@

\begin{table}[h]
  \begin{center}
  \resizebox{1\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline
		\textbf{Critical structure protection}	&	1 Jan - 30 Apr	&	1 Jan - 31 Dec	&	1 Jan - 31 Dec &	1 Jan - 31 Dec &	1 Jan - 31 Dec\\
                	&	2020	&	2019	&	2018 &	2017 &	2016\\
		\hline									
		Relevant attacks	&	282	&	1.181	&	459	&	1.032 & 844	\\
		Issued alerts	&	24.824	&	82.484	&	80.777	&	31.524 & 6.721	\\
		Initiated investigations 	&	34	&	155	&	74	&	72 & 70	\\
		People arrested	&	0	&	3	&	1	&	3	& 3\\
		People reported	&	0	&	117	&	14	&	1.316 &	1.226	\\
		Perquisitions	&	n.d.	&	n.d.	&	n.d.	&	73	& 58 \\
		\hline	
		Initiated investigations	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
		People arrested out ofreported &	0\%	&	2,5\%	&	7,14\%	&	0,23\%	&	0,24\%\\						
		\hline	
	\end{tabular}
	}
    \end{center}

@@

\begin{table}[h]
  \begin{center}
	\begin{tabular}{|r|r|r|r|}									
		\hline									
		\textbf{Cyber-terrorism}	&	1 Jan - 30 Apr 2020	&	1 Jan - 31 Dec 2019	&	1 Jan - 31 Dec 2018\\
		\hline									
		Monitored web space	&	\euro 11.962	&	\euro 36.377	&	\euro 36.000\\
		Removed contents 	&	n.a.\	&	n.a.\	&	250\\
		\hline	
	\end{tabular}
        \end{center}
	\caption{Data of cyber terrorism collected by Italian Postal and Communications Police~\cite{oad20}}									
	\label{tab:table3}									
\end{table}

@@

The OAD 2020 survey highlights the presence of cyberattacks of various types in Italy, all technically of an high complexity and sophistication, with a slight increase 
compared to the trend that emerged in the twelve years of the previous OAD surveys, but still below the pick of 2018. Despite the proliferation of cyberattacks
related to the Covid-19 pandemic, in the first four months of 2020 the widespread of attacks is at a level similar to that of 2019. This will be better verified
with the next OAD 2021. The Italian reality, made up of a very large number of small and tiny organizations, does not make our country very attractive to cybercriminals, but cyber-warfare
and massive attacks represent a growing and serious risk, as has already occurred with the widespread of ransomware on computer systems with no or low
measures of cybersecurity. The OAD 2020 shows a clear improvement and strengthening of digital security measures, both technical and organizational, even if the most modern 
prevention, protection and management systems that use artificial intelligence techniques are still in their infancy among the respondents. The defense measures and techniques 
in use fight the increasingly sophisticated and smart evolution of attacks, but often too late. The high density of vulnerabilities 
requires different approaches, aiming at making all IT systems interconnected to the Internet and safe, by default and by design. But we are 
still a long way from this goal. In order to improve the actual fight against continuous attacks and cybercrime, it is currently necessary to increase 
awareness on cybersecurity and skills at all levels. It is also paramount to support an effective collaboration between police bodies at the international level.

>>>>

% Institutes for affiliations are also joined by \and,
\institute{
  Associazione Italiana Professionisti della Sicurezza Informatica (AIPSI), Milano, Italy
\and
  Italian Information System Security Association International (ISSA)
\and
Digital Attacks Observatory (OAD) Team, Italy\\
\email{m.bozzetti@aipsi.org}
\and
   University of Verona, Verona, Italy\\
   \email{\{luca.olivieri,fausto.spoto\}@univr.it}\\
 }

@@

The \textit{Digital Attacks Observatory} (OAD)~\cite{oadweb} is the only independent online survey in Italy about intentional digital attacks on IT systems of companies and public
bodies. It is implemented in collaboration with the Italian postal and telecommunication police.
The last OAD 2020 was the twelfth year of consecutive surveys on cyberattacks in Italy.
The OAD survey is not directed at a predefined set of respondents, but allows potential interested
parties to have full and free access to an online questionnaire, in a totally anonymous way.
The survey has collected responses from 310 enterprises in 2020, in comparison
to 296 and 2019 and 269 in 2018.
Thanks to the number of responses collected and their balanced distribution between companies and public bodies of different size and belonging to different economic
sectors, we can provide now a specific picture of the cyberattacks in Italy.

@@

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{pictures/fig2.png}
  \caption{Trend of the reported attacks from the beginning of the OAD survey
    in 2008~\cite{oad20}. This comparison is only valid as a trend and not as a statistical
    measure, since the respondents are different from year to year. The dotted red line shows
    successive waves: after a relative peak of attacks, companies react and strengthen
    their cybersecurity measures, so that, the following year, the trend dicreases.}
  \label{fig:2}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.52]{pictures/fig3.png}
	\end{center}
  \caption{Comparison between cyberattacks occurred in 2019 and in the first quarter of 2020,
    wrt.\ to the organization dimension~\cite{oad20}. The dotted line links
    the unreported attacks and shows that they decrease while moving from tiny to big organizations,
    in terms of employees.}
  \label{fig:3}
\end{figure}

The chart in Fig.~\ref{fig:2} analyzes the trend of cyberattaks.
The figure shows that, from 2007 to 2015, in average, about 40\% of the respondents
reported cyberattacks (leftmost full red line). Since 2015, the trend increased to 45\%
of the respondents (rightmost full red line). 
In 2018, OAD reached the peak of reported attacks and, for the first time, the percent of
reported attacks became larger than that percent of companies that did not report any attack.
After the peak 
in 2018, year 2019 has a lower figure and the same occurs in the first quarter 2020.
The trend for full 2020 will be discovered with the currently ongoing OAD survey.
This view, over several years, does not show significant changes, not even inside
the pandemic period, but shows only a slight increase in cyberattacks 
in the recent years, not necessarily attributable to the pandemic crisis.

@@

\begin{table}[h]
  \begin{center}
  \resizebox{1\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline
		\textbf{Critical structure protection}	&	1 Jan - 30 Apr	&	1 Jan - 31 Dec	&	1 Jan - 31 Dec &	1 Jan - 31 Dec &	1 Jan - 31 Dec\\
                	&	2020	&	2019	&	2018 &	2017 &	2016\\
		\hline									
		Relevant attacks	&	282	&	1.181	&	459	&	1.032 & 844	\\
		Issued alerts	&	24.824	&	82.484	&	80.777	&	31.524 & 6.721	\\
		Initiated investigations 	&	34	&	155	&	74	&	72 & 70	\\
		People arrested	&	0	&	3	&	1	&	3	& 3\\
		People reported	&	0	&	117	&	14	&	1.316 &	1.226	\\
		Perquisitions	&	n.d.	&	n.d.	&	n.d.	&	73	& 58 \\
		\hline	
		Initiated investigations	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
		People arrested out of reported &	0\%	&	2,5\%	&	7,14\%	&	0,23\%	&	0,24\%\\						
		\hline	
	\end{tabular}
	}
    \end{center}

@@

\begin{table}[h]
  \begin{center}
	\begin{tabular}{|r|r|r|r|}									
		\hline									
		\textbf{Cyber-terrorism}	&	1 Jan - 30 Apr 2020	&	1 Jan - 31 Dec 2019	&	1 Jan - 31 Dec 2018\\
		\hline									
		Monitored web space	&	\euro 11.962	&	\euro 36.377	&	\euro 36.000\\
		Removed contents 	&	n.a.\	&	n.a.\	&	250\\
		\hline	
	\end{tabular}
        \end{center}
	\caption{Data of cyber terrorism collected by Italian postal and communication Police~\cite{oad20}.}									
	\label{tab:table3}									
\end{table}

@@

The OAD 2020 survey highlights the presence of cyberattacks of various types in Italy, all technically of an high complexity and sophistication, with a slight increase 
compared to the trend that emerged in the twelve years of the previous OAD surveys, but still below the peak of 2018. Despite the proliferation of cyberattacks
related to the Covid-19 pandemic, in the first four months of 2020 the widespread of attacks is at a level similar to that of 2019. This will be better verified
with the next OAD 2021. The Italian reality, made up of a very large number of small and tiny organizations, does not make our country very attractive to cybercriminals, but cyber-warfare
and massive attacks represent a growing and serious risk, as has already occurred with the widespread of ransomware on computer systems with no or low
measures of cybersecurity. The OAD 2020 shows a clear improvement and strengthening of digital security measures, both technical and organizational, even if the most modern 
prevention, protection and management systems that use artificial intelligence techniques are still in their infancy among the respondents. The defense measures and techniques 
in use fight the increasingly sophisticated and smart evolution of attacks, but often too late. The high density of vulnerabilities 
requires different approaches, aiming at making all IT systems interconnected to the Internet and safe, by default and by design. But we are 
still a long way from this goal. In order to improve the actual fight against continuous attacks and cybercrime, it is currently necessary to increase 
awareness on cybersecurity and skills at all levels. It is also paramount to support an effective collaboration between police bodies at the international level.
